Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves

### 2025-07-24 ms/ms.tex
editing main text

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
    \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Developmental plasticity to high light intensity amplifies the
advantage of amphistomous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are common among herbaceous plants growing in
sunny habitats, including many crop species. This distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 14 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plant grown
under high light intensity benefit more in terms of CO\(_2\) assimilated
for a given stomatal conductance than plants grown under low light,
regardless of measurement light intensity or native light habitat.
Developmental plasticity in leaf thickness and/or density, an adaptive
response to optimize light interception, likely affects CO\(_2\)
diffusion within the leaf, altering amphistomy advantage as a byproduct.
Developmental plasticity may therefore play an important and previously
unnoticed role in explaining the adaptive significance of amphistomy and
the distribution of amphistomatous leaves in nature.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and permitted
vascular plants to grow tall on land by enabling access to CO\(_2\) for
photosynthetic carbon assimilation while preventing hydraulic failure in
variable environments (\emph{1}--\emph{3}). Optimal stomatal function
depends on both dynamic changes in aperture on the scale of minutes to
hours, as well as static anatomy determined by developmental plasticity
and constitutive genetic differences (\emph{4}--\emph{6}). Understanding
how stomata respond to environmental change over daily, developmental,
and evolutionary time is important for understanding adaptation
(\emph{1}, \emph{7}--\emph{12}), predicting paleoclimate from fossil
cuticles (\emph{13}--\emph{15}), and improving crops (\emph{16}).
Stomatal function contributes to global carbon and water cycles
(\emph{17}) and therefore predicting future climate (\emph{18}).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships
(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
adaxial stomata whose aperture can be independently regulated
(\emph{27}--\emph{31}) to control gas exchange through each surface. All
else being equal, gas exchange through stomata on both surfaces
increases CO\(_2\) supply to chloroplasts by providing a second parallel
pathway through leaf intercellular airspaces, enhancing photosynthesis
(\emph{20}, \emph{32}). The extent to which amphistomy increases
CO\(_2\) supply depends on resistance to diffusion in intercellular
airspaces. This resistance can be low in thin, porous, amphistomatous
leaves (\emph{28}, \emph{33}), but may be more substantial in thick,
dense, hypostomatous leaves (\emph{34}). Amphistomatous leaves also lose
more water through evaporation because of a second boundary layer
conductance (\emph{35}), but the additional carbon gain should be enough
to offset this cost in most realistic scenarios (\emph{36}).

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, herbaceous plants in open, high light habitats tend
to have have amphistomatous leaves (\emph{22}, \emph{39}--\emph{44}).
Most other leaves, except those from aquatic habitats, are
hypostomatous, producing stomata only on the lower, abaxial surface.
Even resupinate leaves develop stomata on the lower, albeit adaxial
surface (\emph{45}), suggesting that leaf orientation (lower vs.~upper)
rather than leaf polarity (abaxial vs.~adaxial) is causal. The
covariation between stomatal density ratio and light habitat is both
qualitative and quantitative. A higher proportion of sun leaves are
amphistomatous (\emph{39}) and the proportion of stomata on the upper,
adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
why high light intensity favors amphistomatous dorsivental leaves is an
important first step toward understanding variation in stomatal density
ratio and leaf structure-function relationships more generally.

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers; the additional water loss
through a second boundary layer is typically small (\emph{35}). We
quantify this benefit as the amphistomy advantage (\aax{}), the
log-response ratio of photosynthesis in an amphistomatous leaf compared
to an otherwise identical pseudohypostomatous leaf (\emph{20},
\emph{46}). Why would \aax{} be greater in sun than shade? We consider
three nonmutually exclusive hypotheses that we classify as
`acclimatory', `plastic', and `constitutive' (Fig.~\ref{fig-concept}).

\emph{Acclimatory hypothesis}: Photosynthetic induction to high light
intensity typically involves increases in total leaf stomatal
conductance (increased CO\(_2\) supply), the concentration of active
Rubisco, and electron transport capacity (increased CO\(_2\) demand). A
one-dimensional circuit model using the Farquhar-von Caemmerer-Berry
biochemical model of C\(_3\) photosynthesis (\emph{47}) shows that both
increased stomatal conductance and Rubisco activity should increase
\aax{}, all else being equal (Supporting Information). If the
acclimatory hypothesis is correct, we predict that
\(\aahigh{} > \aalow{}\) for all populations regardless of native
habitat or growth environment.

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions
(\emph{48}). Plastic responses are likely adaptations to optimize
photosynthesis at different light intensities in variable environments
(\emph{49}). Plastic changes in leaf anatomy and biochemistry could
modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
which are associated with high leaf mass per area (LMA), will have lower
\(g_\mathrm{ias}\); leaves with increased total stomatal density and
photosynthetic capacity have greater potential CO\(_2\) supply and
demand. Under the plastic hypothesis, we predict that
\(\aasun{} > \aashade{}\) for all populations and light intensities.
Secondarily, \(\aasun{}\) and \gmaxsun{} should also be positively
associated with native light habitat if genotypes adapted to sunny, open
habitats if they can express a phenotype best adapted to that
environment when leaves develop under high light intensity.

\emph{Constitutive hypothesis}: In environments that are relatively
constant or where environmental change cannot be anticipated by a
reliable cue, natural selection will favor constitutive expression of
optimal phenotypes. We therefore predict genotypes from more sunny, open
habitats will have consistently greater \aax under also measurement and
growth light intensities. For herbaceous plants, light intensity is
largely a function of the tree canopy (\emph{50}). Herbs growing in the
open will regularly experience high light intensity; herbs growing under
a forest canopy will often experience low light intensity.

@@

The primary directional predictions for each hypothesis are summarized
in Table~\ref{tbl-hypotheses}; detailed predictions for results that
would indicate support for multiple hypotheses are in
Table~\ref{tbl-predictions}.

@@

\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
Each point is the estimated population-level LMA (\(x\)-axis) and \aax{}
(\(y\)-axis) for each population grown under high light intensity
(\ppfdequals{2000}) and low light intensity (\ppfdequals{150}). The
points are colored by growth light treatment. The solid line is the
posterior median of the linear regression of individual-level LMA on
\(\aahigh\) with the shaded region showing the 95\% confidence ribbon;
the dashed line is the same for \(\aalow\). Confidence intervals for
each population estimate and individual-level estimates are omitted for
visual clarity.}

@@

We modeled \logA as a quadratic function of \loggsw for each leaf,
measured at low and high light intensity, in both amphistomatous
(untreated) and pseudohypostomatous (propafilm treatment) states.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Plasticity and adaptation to high light intensity amplify the
advantage of amphistomous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}

@@

The presence of stomata on both leaf surfaces (amphistomy) can increase
photosynthesis in C\(_3\) plants by reducing the path length for
CO\(_2\) diffusion between substomatal cavities and chloroplasts.
Amphistomatous leaves are common among herbaceous plants growing in
sunny habitats, including many crop species. This distribution of
amphistomatous leaves in nature may result from an increased
photosynthetic benefit of amphistomy under high light intensity, either
because of acclimatory, plastic, or constitutive variation in CO\(_2\)
supply or demand. We used a recently developed method to quantify
amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
relative to an otherwise identical hypostomatous leaf, in 29 diverse
populations representing 14 species of wild tomatoes (\emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
grown under high light intensity benefit more in terms of CO\(_2\)
assimilated for a given stomatal conductance than plants grown under low
light, regardless of measurement light intensity. Furthermore,
populations native to open habitats tend to benefit more from amphistomy
compared to those from more closed habitats. Developmental plasticity
and adaptation of leaves to high light intensity likely play a combined
role in explaining the adaptive significance and distribution of
amphistomy in nature.

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
changes in aperture on the scale of minutes to hours, as well as static
anatomy determined by developmental plasticity and constitutive genetic
differences (\emph{4}--\emph{6}). Understanding how stomata respond to
environmental change over daily, developmental, and evolutionary time is
important for understanding adaptation (\emph{1}, \emph{7}--\emph{12}),
predicting paleoclimate from fossil cuticles (\emph{13}--\emph{15}), and
improving crops (\emph{16}). Stomatal function contributes to global
carbon and water cycles (\emph{17}) and therefore predicting future
climate (\emph{18}).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships
(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
adaxial stomata whose aperture can be independently regulated
(\emph{27}--\emph{31}) to control gas exchange through each surface. All
else being equal, simultaneous gas exchange through stomata on both
surfaces increases CO\(_2\) supply to chloroplasts by providing a second
parallel pathway through leaf intercellular airspaces, enhancing
photosynthesis (\emph{20}, \emph{32}). The extent to which amphistomy
increases CO\(_2\) supply depends on resistance to diffusion in
intercellular airspaces. This resistance can be low in thin, porous,
amphistomatous leaves (\emph{28}, \emph{33}), but may be more
substantial in thick, dense, hypostomatous leaves (\emph{34}). We refer
to the intercellular airspace conductance (\gias), the inverse of
resistance. Amphistomatous leaves also lose more water through
evaporation because of a second boundary layer conductance (\emph{35}),
but the additional carbon gain should be enough to offset this cost in
most realistic scenarios (\emph{36}).

The paradoxical fact is that, despite the photosynthetic benefit, most
leaves are not amphistomatous. Many vertically oriented and/or
isobilateral leaves are amphistomatous (\emph{25}). But among
dorsiventral leaves, it is primarily herbaceous plants in open, high
light habitats that tend to have amphistomatous leaves (\emph{22},
\emph{39}--\emph{44}). Most other leaves, except those from aquatic
habitats, are hypostomatous, producing stomata only on the lower,
abaxial surface. Even resupinate leaves develop stomata on the lower,
albeit adaxial surface (\emph{45}), suggesting that leaf orientation
(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
causal. The covariation between stomatal density ratio and light habitat
is both qualitative and quantitative. A higher proportion of sun leaves
are amphistomatous (\emph{39}) and the proportion of stomata on the
upper, adaxial surface increases with light (\emph{42}, \emph{43}).
Resolving why high light intensity favors amphistomatous dorsivental
leaves is an important first step toward understanding variation in
stomatal density ratio and leaf structure-function relationships more
generally.

The overarching hypothesis is that leaves with greater stomatal density
ratio are more common in open, sunny habitats because they increase
photosynthesis most in those circumstances. An amphistomatous leaf
increases photosynthetic carbon gain compared to an otherwise identical
hypostomatous leaf by increasing conductance through the leaf
intercellular airspaces and boundary layers. We quantify this benefit as
the amphistomy advantage (\aax{}), the log-response ratio of
photosynthesis in an amphistomatous leaf compared to an otherwise
identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
\aax{} be greater in sun than shade? We consider three nonmutually
exclusive hypotheses that we classify as `acclimation', `plasticity',
and `constitutive' (Fig.~\ref{fig-concept}).

\emph{Acclimation hypothesis}: Photosynthetic induction to high light
intensity typically involves increases in total leaf stomatal
conductance (increased CO\(_2\) supply), the concentration of active
Rubisco, and electron transport capacity (increased CO\(_2\) demand). If
the acclimation hypothesis is correct, we predict that
\(\aahigh{} > \aalow{}\) for all populations regardless of native
habitat or growth environment.

\emph{Plasticity hypothesis}: Individuals of the same genotype often
develop dramatically different leaves in sun and shade conditions
(\emph{47}). Plastic responses are likely adaptations to optimize
photosynthesis at different light intensities in variable environments
(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
which are associated with high leaf mass per area (LMA), will have lower
\gias; leaves with increased total stomatal density and photosynthetic
capacity have greater potential CO\(_2\) supply and demand. Under the
plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
populations and light intensities.

\emph{Constitutive hypothesis}: In environments that are relatively
constant or where environmental change cannot be anticipated by a
reliable cue, natural selection will favor constitutive expression of
optimal phenotypes. We therefore predict genotypes from more sunny, open
habitats will have consistently greater \aax{} under all measurement and
growth light intensities. For herbaceous plants, light intensity is
largely a function of the tree canopy (\emph{49}). Herbs growing in the
open will regularly experience high light intensity; herbs growing under
a forest canopy will often experience low light intensity.

@@

The primary directional predictions for each hypothesis are summarized
in Table~\ref{tbl-hypotheses}; detailed predictions for results that
would indicate simultaneous support for multiple hypotheses are in
Table~\ref{tbl-predictions}.

@@

\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
Each point is the estimated population-level LMA (\(x\)-axis) and \aax{}
(\(y\)-axis) for each population grown under high light intensity
(\ppfdequals{2000}) and low light intensity (\ppfdequals{150}). The
points are colored by growth light treatment. The solid line is the
posterior median of the linear regression of individual-level LMA on
\(\aahigh\) with the shaded region showing the 95\% confidence ribbon;
the dashed line is the same for \(\aalow\). Confidence intervals for
each population estimate were omitted for visual clarity.}

@@

We modeled \logA{} as a quadratic function of \loggsw{} for each leaf,
measured at low and high light intensity, in both amphistomatous
(untreated) and pseudohypostomatous (propafilm treatment) states.
